fMRI by caveat

### 2023-03-19 paper.qmd
up to number 11

@@

A more trivial point, perhaps, but important nonetheless. The nice blobs we see in the newspaper indicating the 'love', 'jealousy', or 'political affiliation' region in the brain aren't images of _activation_. They are a value, often a T-statistic, that relates to the certainty we have when we conclude some difference between conditions exist. For example, that 'love' blob might reflect our confidence when we conclude that the BOLD response in this region is different when looking at photos of loved ones vs photos of people we don't know. We need to keep in mind that how to regard this result requires knowing more about 1. the experiment, and 2. the analysis that delivered this result. Remember, there are [lies, damned lies, and statistics](https://en.wikipedia.org/wiki/Lies,_damned_lies,_and_statistics).

@@

::: {.column-margin}
![Spin-Echo animation from Wikipedia](https://mriquestions.com/uploads/3/4/5/7/34572113/_7041819_orig.gif){fig-alt="Spin-Echo animation"}
:::

### Caveat IX: The shape of the BOLD response differs a lot across voxels, individuals, and age. {#sec-c9}

@@

- voxels containing more and/or larger veins will have a stronger BOLD response with a slower rise time and a later peak [@10.1038/jcbfm.2011.57].
- for ultra-high resolution fMRI, this means that the voxels closer to the pial surface have a 2-fold stronger signal than voxels closer to the white-matter gray-matter boundary [@10.1038/s41592-020-0941-6].
- the original HRF shapes were derived from 1.5-Tesla measurements [@10.1523/jneurosci.16-13-04207.1996]. Present-day 7-Tesla measurements show a faster response because they inherently weigh signals from smaller veins more strongly, and these are faster to respond whenever neurons activate.
- we only have good estimates of the HRF shape in primary brain regions, such as primary motor cortex or primary visual cortex. This is because only when we can provide an 'impulse' (a very strong and very brief stimulus) can we estimate the shape of the impulse-response to this input event. For many other brain regions we simply don't know how to activate neurons briefly and strongly, but for motor or visual cortex that can be done by finger-tapping or presenting a flickering checkerboard stimulus, respectively.

@@

The use of the GLM depends on the assumption that the relation between the BOLD response and the thing of interest (neural responses) is a linear time-invariant (LTI) system. This means that doubling the strength of a neural signal will show a double amplitude response in the BOLD response. It also means that the response to a two-second event is equal to the sum of two one-second event responses, the second having been shifted by 1 second. Using these assumptions we can now use convolution with the impulse-response of the system (the HRF) to predict BOLD responses driven by specific events. Also, this means that we can gauge the strength of neural responses by looking at the $\beta$ weights that roll out of a GLM. It's important to realize that this is an _assumption_ that we cannot fully verify. It's clear that as a first pass it holds, though, but as opposed to simple electrical wirings, the complicated biological machinery driving the BOLD response is likely to show nonlinearities that break the LTI assumption. Just check out the examples above in Caveat @sec-c5 to convince yourself that there are limits to the LTI assumption [@10.1038/nature07664].

### Caveat XI: GLM statistics depend on specific assumptions. {#sec-c11}

The statistics we do on top of our GLMs assume independent, identically distributed samples - an assumption that is very common in more statistical applications. However, the noise that ends up as residuals actually has strong temporal autocorrelation. This means that the covariance structure of the noise doesn't follow an identity matrix, and that we should estimate this structure and use it to correct our statistics. If we do not take into account this correlational structure in our noise, we will overestimate the degrees of freedom that go into our T-statistic calculations, greatly inflating them and causing a lot of false positives. Similarly, cluster-based statistical procedures often use assumptions regarding the spatial autocorrelation of noise. If we assume such structure to be more narrow than it actually is, we run into problems [@10.1073/pnas.1602413113]. These problems, as in @sec-c10, are that we will be too lenient and produce false-positive results.

>>>>

Regardless of caveats, a great resource for the concepts underpinning MR imaging is the [mriquestions website](https://mriquestions.com).

@@

A more trivial point, perhaps, but important nonetheless. The nice blobs we see in the newspaper indicating the 'love', 'jealousy', or 'political affiliation' region in the brain aren't images of _activation_. They are a value, often a T-statistic, that relates to the certainty we have when we conclude some difference between conditions exist. For example, that 'love' blob might reflect our confidence when we conclude that the BOLD response in this region is different when looking at photos of loved ones vs photos of people we don't know (usually this is a 'baseline' with similar low-level stimulus features, faces, in this case). We need to keep in mind that how to interpret this result requires knowing more about 1. the experiment, and 2. the analysis that delivered this result. Remember, there are [lies, damned lies, and statistics](https://en.wikipedia.org/wiki/Lies,_damned_lies,_and_statistics).

@@

::: {.column-margin}
![Spin-Echo animation from Wikipedia](figs/_7041819_orig.gif){fig-alt="Spin-Echo animation"}
:::

### Caveat IX: The shape of the Haemodynamic Response Function differs a lot across voxels, individuals, and age. {#sec-c9}

@@

- voxels containing more and/or larger veins will have a stronger BOLD response with a slower rise time and a later peak [@10.1038/jcbfm.2011.57; @10.1038/s41592-020-0941-6].
- for ultra-high resolution fMRI, this means that the voxels closer to the pial surface have a 2-fold stronger signal than voxels closer to the white-matter gray-matter boundary[@10.1016/j.neuroimage.2020.117683].
- the original HRF shapes were derived from 1.5-Tesla measurements [@10.1523/jneurosci.16-13-04207.1996]. Present-day 7-Tesla measurements show a faster response because they inherently weigh signals from smaller veins more strongly, and these are faster to respond whenever neurons activate.
- we only have good estimates of the HRF shape in primary brain regions, such as primary motor cortex or primary visual cortex. This is because only when we can provide an 'impulse' (a very strong and very brief stimulus) can we estimate the shape of the impulse-response to this input event. For many other brain regions we simply don't know how to activate neurons briefly and strongly, but for motor or visual cortex that can be done by finger-tapping or presenting a flickering checkerboard stimulus, respectively.

@@

The use of the GLM depends on the assumption that the relation between the BOLD response and the thing of interest (neural responses) is a linear time-invariant (LTI) system. This means that doubling the strength of a neural signal will show a double amplitude response in the BOLD response. It also means that the response to a two-second event is equal to the sum of two one-second event responses, the second having been shifted by 1 second. These two represent the linear (mulitiplication and addition) part of the term. Lastly, we assume that these relations are stable over time (time-invariant). Using these assumptions we can now use convolution with the impulse-response of the system (the HRF) to predict BOLD responses driven by specific events. Also, this means that we can gauge the strength of neural responses by looking at the $\beta$ weights that roll out of a GLM. It's important to realize that seeing the relation between neural activations and BOLD response as an LTI is an _assumption_ that we cannot fully verify. It's clear that as a first pass it holds[@10.1523/jneurosci.16-13-04207.1996], though. However, as opposed to simple electrical wirings, the complicated biological machinery driving the BOLD response is likely to show nonlinearities that break the LTI assumption. Just check out the examples above in Caveat @sec-c5 to convince yourself that there are limits to the LTI assumption [@10.1038/nature07664].

### Caveat XI: GLM statistics depend on specific assumptions, that are not always met. {#sec-c11}

The statistics we do on top of our GLMs assume independent, identically distributed samples - an assumption that is very common in more statistical applications. However, the noise that ends up as residuals actually has strong autocorrelations, spatial and temporal.

In the temporal case, this means that the covariance structure of the noise doesn't follow an identity matrix, and that we should estimate this structure and use it to correct our statistics (i.e. move from an [ordinary least squares](https://en.wikipedia.org/wiki/Ordinary_least_squares) approach to a [generalized least squares](https://en.wikipedia.org/wiki/Generalized_least_squares) approach). If we do not take into account this correlational structure in our noise, we will overestimate the degrees of freedom that go into our T-statistic calculations, greatly inflating them and causing a lot of false positives.

Similarly, cluster-based statistical procedures often use assumptions regarding the spatial autocorrelation of noise. If we assume such structure to be more narrow than it actually is, we run into problems [@10.1073/pnas.1602413113]. These problems, as in @sec-c10, are that we will be too lenient and produce false-positive results.

### Caveat XII: When performing a decoding analysis, you're leaving the hard work to a black-box algorithm. {#sec-c12}

Decoding from a pattern of BOLD responses across voxels which stimulus is shown or experienced by a participant is a sort of 'mind reading'. There are a lot of papers that perform this type of analysis and there are some awesome findings among them. But it's important to think deeply about what we're doing here. As argued in this nice opinion paper[@10.1016/j.tics.2015.07.005], leaving the inferential ....
